### 2010-03-13 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: merge Conor's comments in. (first pass)

Finally, we define the rules governing judgemental equality in
Figure~\ref{fig:judgemental-equality}. As one would expect,
judgemental equality respects symmetry, reflexivity, and
transitivity. Further, it also encompasses the $\beta$-rules for
function application and pairs. Finally, equality under lambda is
dealt with by the $\xi$-rule.

Crucially, being judgemental, this presentation is agnostic in the
notion of equality actually implemented. Indeed, our typing and
equality judgements do not rely on a ``propositional'' equality. This
freedom is a key point in favour of the wide applicability of our
proposal. Obviously, when implementing our technology in an existing
type-theory, some opportunities arise. We will present some of them
along the course of the paper.

@@

However, using encodings is impractical. In the case of finite sets,
for instance, we would like to name the elements of the sets. Then, we
need to be able to manipulate these elements by their name, instead of
their encoding. While we are able to give names to encodings, it is
extremely tedious to map the encodings back to a name. Whereas these
objects have a structure, the structure is lost during the encoding,
when they become anonymous inhabitants of a $Π$ or $Σ$-type.

@@

The $\Enum$ type comes with further machinery. Indeed, to manipulate
the $\EnumU$ and $\EnumT{}$ objects, we need to equip our type theory
with an appropriate \emph{elimination principle}. A first step
consists in defining the $\spi{}{}$ operator:

\[
%% spi
\spi{}{} : \PITEL{\V{e}}{\EnumU}
           \PITEL{\V{P}}{\EnumT{e} \To \Set} \To \Set 
\]

@@

\[
%% switch
\begin{array}{@{}ll}
\F{switch} : & \PITEL{\V{e}}{\EnumU}
               \PITEL{\V{P}}{\EnumT{e} \To \Set} \\
             & \PITEL{\V{b}}{\spi{e}{P}}
               \PITEL{\V{x}}{\EnumT{e}} \To P~x
\end{array}
\]

@@

As we have seen, finite sets are not just a way of making
data. Through the $\spi{}{}$ and $\F{switch}$ operators, they also
come with a notion of finite function space. However, we had to
extract that intuition from the type, by a careful reading. At a first
glance, this seems to contradict our argument in favour of types for
coding representation. Here, we are overflown by low-level, very
precise type information.

However, our situation is significantly different from the one we
faced with encoded data: while we were suffering from a crucial lack
of information, we are here facing too much information, hence loosing
focus. This is a general issue with the usage of codes, as they convey
much more information than what the user is used to see.

As we will see in the following section, there exists a cure to this
problem. In a nutshell, instead of being overflown by typing
information, we will put it at work, automatically. The consequence is
that, in such system, working with codes is \emph{practical}: one
should not be worried by information overload, but how to use it as
much as possible. Having mastered this difficulty, we should not be
afraid by code for practical purposes.

@@

On the other hand, in the \emph{type checking} phase, terms are
\emph{pushed} into types. We are handed a type together with a term,
our task consists in checking that the term is indeed of this type. We
are therefore allowed, and actually encouraged, to use the information
provided by the type. Therefore, we can relax our requirements on the
term. A typical example is the lambda:

\[
\Rule{\Gamma ; \xS \Vdash \propag{\push{t}{T\: x}}
                                 {t'}}
     {\Gamma \Vdash \propag{\push{\LAM{\x} t}{\PiTy{S}{T}}}
                           {\PLAM{\x}{S} t'}} 
\]

In our formal presentation of the type theory, we were forced to
annotate the lambda by the type of its domain. In a bidirectional
setting, the domain is provided by the $\Pi$ type we are pushing the
lambda into. Hence we do not need to annotate the lambda.

@@

Dually to type synthesis, type checking judgements
(Fig.\ref{fig:type-checking}) act as \emph{sinks}. From an expression
$\exprIn$ and a type propagated down to it, they elaborate a low-level
term. This is precisely the availability of the type we are pushed
into that makes the power of this technique. Its benefit on lambda
abstraction and pairs is well-known. Let us explore the consequences
on our coding scheme for finite sets.

@@

The abstraction we have developed in this section are entirely
transparent to the user. In the case of finite function space, the
user has never to explicitly \emph{call} the $\F{switch}$
eliminator. Nor will she \emph{be presented} the elaborated term, only
the high-level . Bidirectional type-checking and type-directed
pretty-printing seal the low-level type theory behind the high-level
expressions.

@@

This example teaches us that we should not be afraid by codes, as far
as type information is available. We have shown how to rationalise
this information in a formal presentation. Hence, we have proved that
programming with such objects is practical. As we introduce more codes
in our theory, we will show how to extend the framework we have
developed so far.

>>>>

Finally, we define the rules governing judgemental equality in
Figure~\ref{fig:judgemental-equality}. As one would expect,
judgemental equality respects symmetry, reflexivity, and
transitivity. Further, it encompasses the $\beta$-rules for function
application and pairs. Finally, it respects structural equality,
including equality under lambda ($\xi$-rule).

Crucially, being judgemental, this presentation is agnostic in the
notion of equality actually implemented. Indeed, our typing and
equality judgements do not rely on a ``propositional'' equality. This
freedom is a key point in favour of the wide applicability of our
proposal. Obviously, when implementing our technology in an existing
type-theory, some opportunities arise. We will present some of them
along the course of the paper. \note{It is just a specification. We
  expect the propositional theory to provide a decidable machinery to
  implement it.}

\note{plus structural rules, including lambda ($\xi$-rule)}

@@

However, using encodings is impractical. In the case of finite sets,
for instance, we would like to name the elements of the sets. Then, we
need to be able to manipulate these elements by their name, instead of
their encoding. While we are able to give names to encodings, it is
extremely tedious to map the encodings back to a name. Whereas these
objects have a structure, the structure is lost during the encoding,
when they become anonymous inhabitants of a $\Pi$ or $\Sigma$-type.

@@

The type of finite sets comes with further machinery. Indeed, to
manipulate the $\EnumU$ and $\EnumT{}$ objects, we need to equip our
type theory with an appropriate \emph{elimination principle}. A first
step consists in defining the $\spi{}{}$ operator:

\[\stk{
%% spi
\spi{}{} : \PITEL{\V{e}}{\EnumU}
           \PITEL{\V{P}}{\EnumT{e} \To \Set} \To \Set \\
\begin{array}{@{}ll@{\:\mapsto\:\:}l}
\spi{\NilE}{& P}        & \Unit \\
\spi{(\ConsE{t}{e})}{& P} & \TIMES{P\: \Ze}{\spi{e}{(\LAM{x} P\: (\Su{x}))}}
\end{array}
}\]

\note{Needs to be defined.}

@@

\[\stk{
%% switch
\begin{array}{@{}ll}
\F{switch} : & \PITEL{\V{e}}{\EnumU}
               \PITEL{\V{P}}{\EnumT{e} \To \Set} \\
             & \PITEL{\V{b}}{\spi{e}{P}}
               \PITEL{\V{x}}{\EnumT{e}} \To P~x
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\switch{(\ConsE{t}{e})}{P}{b}{\Ze}      & \fst{b} \\
\switch{(\ConsE{t}{e})}{P}{b}{(\Su{x})} & \switch{e}{(\LAM{x} P
  (\Su{x}))}{(\snd{b})}{x}
\end{array}
}\]

@@

As we have seen, finite sets are not just a way of making
data. Through the $\spi{}{}$ and $\F{switch}$ operators, they also
come with a notion of finite function space. However, we had to
extract that intuition from the type, by a careful reading. This seems
to contradict our argument in favour of types for coding
representation. Here, we are overflown by low-level, very precise type
information.

However, our situation is significantly different from the one we
faced with encoded data: while we were suffering from a crucial lack
of information, we are now facing too much information, hence loosing
focus. This is a general issue with the usage of codes, as they convey
much more information than what the user is willing to see.

As we will see in the following section, there exists a cure to this
problem. In a nutshell, instead of being overflown by typing
information, we will put it at work, automatically. The consequence is
that, in such system, working with codes is \emph{practical}: one
should not be worried by information overload, but how to use it as
much as possible. Therefore, we should not be afraid by using code for
practical purposes.

@@

On the other hand, in the \emph{type checking} phase, terms are
\emph{pushed} into types. We are handed a type together with a term,
our task consists in checking that the term is indeed of this
type. Doing so, we are allowed, and actually encouraged, to use the
information provided by the type. Therefore, we can relax our
requirements on the term. A typical example is the lambda abstraction:

\[
\Rule{\Gamma ; \xS \Vdash \propag{\push{t}{T}}
                                 {t'}}
     {\Gamma \Vdash \propag{\push{\LAM{\x} t}{\PIS{\xS}{T}}}
                           {\PLAM{\x}{S} t'}} 
\]

In our formal presentation of the type theory, we were forced to
annotate lambda abstractions by the type of the domain. In a
bidirectional setting, the domain is provided by the $\Pi$ type we are
pushing the abstraction into. Hence we do not need the annotation.

@@

Dually to type synthesis, type checking judgements
(Fig.\ref{fig:type-checking}) act as \emph{sinks}. From an expression
$\exprIn$ and a type propagated down to it, they elaborate a low-level
term. The strength of this approach comes from the availability of the
type to drive the elaboration of expressions. Its benefit on lambda
abstraction and pairs is well-known. Let us explore the consequences
on our coding scheme for finite sets.

@@

The abstraction we have developed in this section are entirely
transparent to the user. In the case of finite function space, the
user has never to explicitly \emph{call} the $\F{switch}$
eliminator. Nor will she \emph{be presented} the elaborated term, only
the high-level one. Bidirectional type-checking and type-directed
pretty-printing seal the low-level type theory behind the high-level
expressions.

@@

This example teaches us that we should not be afraid by codes, as soon
as type information is available. We have shown how to rationalise
this information in a formal presentation. Hence, we have proved that
programming with such objects is practical. As we introduce more codes
in our theory, we will show how to extend the framework we have
developed so far.
